The Ranked Nearest Neighbour Interchange space

### 2019-05-15 paper.tex
split + cluster thm

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{split_theorem_counterexample}
	\caption{The split $123|45$ is present in $T_1$ and $T_2$, but in none of the trees on this shortest caterpillar path between them.}
	\label{split_theorem_counterexample}
\end{figure}

Since this counterexample clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem.
This path is shown in Figure~\ref{split_theorem_no_counterexample}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{split_theorem_no_counterexample}
	\caption{The Split $123|45$ is present in $T_1$ and $T_2$ and in every tree on this shortest path between them.}
	\label{split_theorem_no_counterexample}
\end{figure}

@@

% The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).
% We might want to consider the Cluster Theorem as well.
% The weak split theorem is stronger than the cluster theorem, since it also makes a statement about some trees that do not share a cluster

%A further and even weaker alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
%One can see that the trees of our counterexample~\ref{split_theorem_counterexample} do not fulfil the assumptions of the cluster theorem.
%
%\begin{conjecture}[Cluster Theorem]
%	For the $\rnni$ graph the following statement holds:
%	if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
%	\label{cluster_theorem}
%\end{conjecture}

@@

% Also mention the exact algortihm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
%Maybe it's worth trying the other algorithm in these trees as well.

@@

\begin{theorem}
	For $n \geq 3$, $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{theorem}

@@

% Is this boundary tight? Idea: there are always caterpillar trees that have this distance (we can already check this computationally, if we were to prove the caterpillar convexity, this follows directly)

@@

\begin{proof}
    For this proof we will give two caterpillar trees $T_1$ and $T_2$ on $n$ taxa with $d_{RNNI}(T_1,T_2) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
    \todo{$d$ or $d_{RNNI}$?}
	Let $T_1 = (( \dots (n,1)2)\dots)n-1)$ and $T_2 = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Lemma~\ref{lemma:caterpillar_subset_convex}, there is a shortest path between $T_1$ and $T_2$ that only consists of caterpillar trees.
    We can compute such a shortest caterpillar path by using a modified version of Bubble Sort.
    \todo{Explain how to use Bubble Sort}
    When running this algorithm on $T_1$ and $T_2$ there are $n-1-i$ swaps of taxon $i$ with its right neighbour.
    Therefore, the total number of these swaps, and therefore $d_{RNNI}(T_1,T_2)$, is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{proof}

\begin{corollary}
	If Conjecture~\ref{conj:caterpillar_diameter} holds, it is $\Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
\end{corollary}

>>>>

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{splitthm_counterexample}
	\caption{The split $123|45$ is present in $T_1$ and $T_2$, but the path at the top is a shortest path where none of the trees contains this split.
    However, this split is maintained on the path at the bottom that is a shortest path as well.}
	\label{splitthm_counterexample}
\end{figure}

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem, which is presented at the bottom of Figure~\ref{splitthm_counterexample}.

@@

\todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}

% The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).

Another alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
Considering clusters rather than splits is more reasonable since two rooted trees have the same set of clusters if, and only if, they are isomorphic~\cite{steel2016}.
The trees $T_1, T_2$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.

\begin{conjecture}[Cluster Theorem]
	For the $\rnni$ graph the following statement holds:
	if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
	\label{cluster_theorem}
\end{conjecture}

@@

% Also mention the exact algortihm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
% Maybe it's worth trying the other algorithm in these trees as well.

@@

\begin{theorem}
	It is $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{theorem}

@@

@@

\begin{proof}
    For this proof we will give two caterpillar trees $T_1$ and $T_2$ on $n$ taxa with $d_{RNNI}(T_1,T_2) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
    \todo{$d$ or $d_{RNNI}$?}
	Let $T_1 = (( \dots (n,1)2)\dots)n-1)$ and $T_2 = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Lemma~\ref{lemma:caterpillar_subset_convex}, there is a shortest path between $T_1$ and $T_2$ that only consists of caterpillar trees.
    We can compute such a shortest caterpillar path by using a modified version of Bubble Sort.
    When running this algorithm on $T_1$ and $T_2$ there are $n-1-i$ swaps of taxon $i$ with its right neighbour.
    Therefore, the total number of these swaps, and therefore $d_{RNNI}(T_1,T_2)$, is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{proof}

\begin{corollary}
It is $\Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
\end{corollary}
